---
title: "Single-term presidents vs multi-term presidents"
output: html_notebook
---
## Did their speeches differ?

In this project, I looked at the presidents' inauguation speeches for those who belong to either Democratic or Republican parties. I compared the presidents who only served one-term to those who served multiple terms, and tried to find if there are any differences.

# Step 1: Install and load packages
```{r}
rm(list=ls())

# check packages that need to be installed.
packages.used=c("readxl","readr","tm", "wordcloud", "RColorBrewer", "dplyr",
                "tidytext","rvest","tibble","qdap","sentimentr",
                "gplots","syuzhet","factoextra","beeswarm","scales",
                "RColorBrewer","RANN","topicmodels","ggplot2","reshape")

packages.needed=setdiff(packages.used, intersect(installed.packages()[,1], packages.used))

# install additional packages
if(length(packages.needed)>0){
  install.packages(packages.needed, dependencies = TRUE,repos='http://cran.us.r-project.org')
}

# load packages
library(readxl)
library(readr)
library(tm)
library(wordcloud)
library(RColorBrewer)
library(dplyr)
library(tidytext)
library(rvest)
library(tibble)
library(qdap)
library(sentimentr)
library(gplots)
library(syuzhet)
library(factoextra)
library(beeswarm)
library(scales)
library(RColorBrewer)
library(RANN)
library(topicmodels)
library(ggplot2)
library(reshape)

```

```{r}
print(R.version)
```

# Step 2: Pre-processing
### Group data - Republicans & Democrats
```{r}
# Group data
# Republicans vs Democrats
inaug_info <- read_excel("../data/InaugurationInfo.xlsx")
inaug_info_GOP <- inaug_info[inaug_info$Party=='Republican',]
inaug_info_DNC <- inaug_info[inaug_info$Party=='Democratic',]
inaug_info <- rbind(inaug_info_GOP,inaug_info_DNC) #Removing those who are not GOP or DNC.

dim(inaug_info_GOP)
dim(inaug_info_DNC)
dim(inaug_info)
```
### Group data - single-term & multi-term presidents
```{r}
# Speeches of presidents who served multiple terms vs one term
multi_term_pres <- inaug_info[inaug_info$Term!='1',]
one_term_pres <- inaug_info[-which(inaug_info$President %in% multi_term_pres$President),]
multi_term_pres <- inaug_info[which(inaug_info$President %in% multi_term_pres$President),]

# handle an exception -- Grover Cleveland
one_term_pres <- one_term_pres[one_term_pres$President != 'Grover Cleveland - I',]
multi_term_pres <- rbind(multi_term_pres,inaug_info[inaug_info$President=='Grover Cleveland - I',])

# Add a column to identify one-term or multi-term president.
one_term_pres$Type <- 'Single-Term'
multi_term_pres$Type <- 'Multi-Term'

# Check and overwrite inaug_info.
dim(one_term_pres)
dim(multi_term_pres)

inaug_info <- rbind(one_term_pres,multi_term_pres)
```
### Add text data & NRC sentiments to the inaug_info file
```{r}
# Pre-processing
folder.path="../data/InauguralSpeeches/"

# Add text to the inaug_info df.
inaug_info$fulltext <- NA

for(i in seq(nrow(inaug_info))){
  original_file = paste0("inaug", as.vector(inaug_info$File[i]),
                         "-",as.vector(inaug_info$Term[i]),".txt")
  text = read_file(file.path(folder.path, original_file))
  inaug_info$fulltext[i]=text
}

inaug_info
```

```{r}
#Create a sentence list with NRC sentiments
sentence.list<-NULL
for(i in 1:nrow(inaug_info)){
  sentences=sent_detect(inaug_info$fulltext[i],endmarks = c("?", ".", "!", "|",";"))
  if(length(sentences)>0){
    emotions=get_nrc_sentiment(sentences)
    word_count=word_count(sentences)
    emotions=diag(1/(word_count+0.01))%*%as.matrix(emotions)
    sentence.list=rbind(sentence.list, 
                        cbind(inaug_info[i,-ncol(inaug_info)],
                              sentences=as.character(sentences), 
                              word_count,
                              emotions,
                              sent_id=1:length(sentences)
                        )
    )
  }
}
sentence.list= sentence.list%>% filter(!is.na(word_count)) 
summary(sentence.list)
```

### Group speech files into directories by term length
```{r}
### Create a dir for multi_term presidents' speeches
dir.create(file.path(folder.path, "multi_term"),showWarnings = FALSE)
for(i in seq(nrow(multi_term_pres))){
  original_file = paste0("inaug", as.vector(multi_term_pres$File[i]),
                    "-",as.vector(multi_term_pres$Term[i]),".txt")
  text = read_file(file.path(folder.path, original_file))
  new_file <- paste0(folder.path, "multi_term/", original_file)
  sink(file = new_file) %>% 
    cat(text)  # write the file
  sink()
}

### Create a dir for single_term presidents' speeches
dir.create(file.path(folder.path, "single_term"),showWarnings = FALSE)
for(i in seq(nrow(one_term_pres))){
  original_file = paste0("inaug", as.vector(one_term_pres$File[i]),
                         "-",as.vector(one_term_pres$Term[i]),".txt")
  text = read_file(file.path(folder.path, original_file))
  new_file <- paste0(folder.path, "single_term/", original_file)
  sink(file = new_file) %>% 
    cat(text)  # write the file
  sink()
}
```

# Step 3 Data Analysis - word cloud & frequent words
First, let's look at what kind of words were used in the speeches, and if there are any significant differences between the two groups (single-term vs multi-term).
```{r}
###############################################################
#### Word cloud & frequent words
###############################################################

single_term_path <- paste0(folder.path, "single_term/")
multi_term_path <- paste0(folder.path, "multi_term/")

single_term_speeches =list.files(path = single_term_path, pattern = "*.txt")
multi_term_speeches=list.files(path = multi_term_path, pattern = "*.txt")
noprex_single_term_sp=substr(single_term_speeches, 6, nchar(single_term_speeches)-4)
noprex_multi_term_sp=substr(multi_term_speeches, 6, nchar(multi_term_speeches)-4)

ff_single_term <- Corpus(DirSource(single_term_path))
ff_multi_term <- Corpus(DirSource(multi_term_path))

cleanupCorpus <- function(ff){
  ff<-tm_map(ff, stripWhitespace)
  ff<-tm_map(ff, content_transformer(tolower))
  ff<-tm_map(ff, removeWords, stopwords("english"))
  ff<-tm_map(ff, removeWords, character(0))
  ff<-tm_map(ff, removePunctuation)
}

ff_single_term = cleanupCorpus(ff_single_term)
ff_multi_term = cleanupCorpus(ff_multi_term)

tdm_single <- TermDocumentMatrix(ff_single_term)
tdm_single_tidy = tidy(tdm_single)
tdm_single_overall=summarise(group_by(tdm_single_tidy, term), sum(count))

tdm_multi <- TermDocumentMatrix(ff_multi_term)
tdm_multi_tidy = tidy(tdm_multi)
tdm_multi_overall=summarise(group_by(tdm_multi_tidy, term), sum(count))

print('single-term presidents\' frequent terms')
findFreqTerms(tdm_single,100)
print('multi-term presidents\' frequent terms')
findFreqTerms(tdm_multi,100)

set.seed(123)
wordcloud(tdm_single_overall$term, tdm_single_overall$`sum(count)`,
          scale=c(4,0.5),
          max.words=50,
          min.freq=1,
          random.order=FALSE,
          rot.per=0.2,
          use.r.layout=T,
          random.color=FALSE,
          colors=brewer.pal(3,"Dark2"))
wordcloud(tdm_multi_overall$term, tdm_single_overall$`sum(count)`,
          scale=c(4,0.5),
          max.words=50,
          min.freq=1,
          random.order=FALSE,
          rot.per=0.2,
          use.r.layout=T,
          random.color=FALSE,
          colors=brewer.pal(3,"Dark2"))
```
At a glance, single-term presidents' words seem more formal (government, country, people, states, etc.), compared to those by multi-term presidents' words (warming, candor, music, wellinformed, hang, etc.).

Did multi-term presidents tend to use more emotionally-charged words? And did single-term presidents tend to be more formal? 

Let's find out.

# Step 4 Data Analysis - sentiment analysis
In this section, I looked at the magnitude and types of emotions contained in each sentence used by two groups. 

Furthermore, I separated multi-term presidents' speeches into two groups, because I wondered if their first speeches are actually similar to those by the single-term presidents, and they may become more relaxed for their 2nd speeches.

```{r}
# Mean of each emotion score by group
single_emo_means=colMeans(sentence.list%>% 
                            filter(Type=="Single-Term")%>%
                            select(anger:trust)>0.01)
multi_emo_means1=colMeans(sentence.list%>%
                            filter(Type=="Multi-Term",Term==1)%>%
                            select(anger:trust)>0.01)
multi_emo_means2=colMeans(sentence.list%>%
                            filter(Type=="Multi-Term",Term!=1)%>%
                            select(anger:trust)>0.01)
# Merge them as a df.
types = c("single-term","multi-term 1st","multi-term 2nd+")
emo_means = data.frame(types,
                       rbind(single_emo_means,multi_emo_means1,multi_emo_means2),
                       row.names=NULL)
emo_means$types <- factor(emo_means$types, as.character(emo_means$types))
emo_means.m = melt(emo_means,id.vars='types')

# Plot a grouped bar chart
ggplot(emo_means.m, aes(variable, value)) +   
  geom_bar(aes(fill = types), position = "dodge", stat="identity")
```

Unlike what I assumed from looking at the Word Clouds, single-term presidents' sentences tend to have higher scores for emotions.

Interestingly, multi-term presidents' first speeches have lower emotion scores, and the scores became higher for the second speeches. Still, they are lower than the scores by single-term presidents.

Let's look at some of those emotional sentences.

```{r}
### What are the emotionally charged sentences?
print("##### Single Term #####")
speech.df=tbl_df(sentence.list)%>%
  filter(Type=="Single-Term", word_count>=5)%>%
  select(President, sentences, anger:trust)
speech.df=as.data.frame(speech.df)
as.character(speech.df$sentences[apply(speech.df[,-c(1:2)], 2, which.max)])

print("##### Multi Term - 1st speech #####")
speech.df=tbl_df(sentence.list)%>%
  filter(Type=="Multi-Term", Term == 1, word_count>=5)%>%
  select(President, sentences, anger:trust)
speech.df=as.data.frame(speech.df)
as.character(speech.df$sentences[apply(speech.df[,-c(1:2)], 2, which.max)])

print("##### Multi Term - 2nd+ speech #####")
speech.df=tbl_df(sentence.list)%>%
  filter(Type=="Multi-Term", Term != 1, word_count>=5)%>%
  select(President, sentences, anger:trust)
speech.df=as.data.frame(speech.df)
as.character(speech.df$sentences[apply(speech.df[,-c(1:2)], 2, which.max)])
```

# Step 5 Data Analysis - topic modeling

In this section, I looked at if there are any differences in topics discussed in the speeches of single-term presidents v.s. multi-term presidents.
```{r}
# prepare snipets
corpus.list=sentence.list[2:(nrow(sentence.list)-1), ]
sentence.pre=sentence.list$sentences[1:(nrow(sentence.list)-2)]
sentence.post=sentence.list$sentences[3:(nrow(sentence.list)-1)]
corpus.list$snipets=paste(sentence.pre, corpus.list$sentences, sentence.post, sep=" ")
rm.rows=(1:nrow(corpus.list))[corpus.list$sent_id==1]
rm.rows=c(rm.rows, rm.rows-1)
corpus.list=corpus.list[-rm.rows, ]

docs <- Corpus(VectorSource(corpus.list$snipets))
set.seed(123)
writeLines(as.character(docs[[sample(1:nrow(corpus.list), 1)]]))
```
```{r}
# clean up
docs = cleanupCorpus(docs) # the function we used for the word cloud corpus
docs = tm_map(docs, removeNumbers)
docs = tm_map(docs,stemDocument)

set.seed(123)
writeLines(as.character(docs[[sample(1:nrow(corpus.list), 1)]]))
```
```{r}
dtm <- DocumentTermMatrix(docs)

#convert rownames to filenames
rownames(dtm)=paste(substr(corpus.list$Type, 1, nchar(corpus.list$Type)-5), 
                    corpus.list$File, corpus.list$Term, 
                    corpus.list$sent_id, sep="_")

rowTotals=apply(dtm, 1, sum) #sum of words in each doc
dtm = dtm[rowTotals>0, ]
corpus.list=corpus.list[rowTotals>0, ]
```

### Let's run LDA with Gibbs sampling
```{r}
#LDA with Gibbs
burnin = 1000
iter = 2000
thin = 200
seed = list(6558,7,631,10009,4921)
nstart = 5
best = TRUE

k=15 #num of topics
set.seed(123)
ldaOut <-LDA(dtm, k, method="Gibbs", control=list(nstart=nstart, 
                                                  seed = seed, best=best,
                                                  burnin = burnin, iter = iter, 
                                                  thin=thin))
```

```{r}
#write out results
#docs to topics
ldaOut.topics <- as.matrix(topics(ldaOut))
table(c(1:k, ldaOut.topics))
dir.create(file.path('output/LDAGibbs'),showWarnings = FALSE)
write.csv(ldaOut.topics,file=paste0("../output/LDAGibbs/",k,"DocsToTopics.csv"))

#top 10 terms in each topic
ldaOut.terms <- as.matrix(terms(ldaOut,10))
write.csv(ldaOut.terms,file=paste0("../output/LDAGibbs/",k,"TopicsToTerms.csv"))

#probabilities associated with each topic assignment
topicProbabilities <- as.data.frame(ldaOut@gamma)
write.csv(topicProbabilities,file=paste0("../output/LDAGibbs/",k,"TopicProbabilities.csv"))

```

```{r}
terms.beta=ldaOut@beta
terms.beta=scale(terms.beta)
ldaOut.terms
```
### Naming the topics
Rather than guessing the names from terms, I combined the first 3 terms to make them the "Topic name".
```{r}
# Create topic labels by combining the first 3 terms
topics.terms= ldaOut.terms
topics.hash = NULL
for (i in 1:k){
  topic.name = paste(topics.terms[1,i],topics.terms[2,i],
             topics.terms[3,i], sep='-')
  topics.hash = c(topics.hash,topic.name)
}
topics.hash # list of topics
colnames(topics.terms) = topic.hash

corpus.list$ldatopic=as.vector(ldaOut.topics)
corpus.list$ldahash=topics.hash[ldaOut.topics]

colnames(topicProbabilities)=topics.hash
corpus.list.df=cbind(corpus.list, topicProbabilities)

```

### Clustering
```{r}
#clustering
par(mar=c(1,1,1,1))
topic.summary=tbl_df(corpus.list.df)%>%
  filter(Type == "Single-Term")%>%
  select(File, 23:37)%>%
  group_by(File)%>%
  summarise_all(funs(mean))
topic.summary=as.data.frame(topic.summary)
rownames(topic.summary)=topic.summary[,1]

topic.plot=1:15

heatmap.2(as.matrix(topic.summary[,topic.plot+1]), 
          scale = "column",
          col = rich.colors(100),
          cexRow = 0.9, cexCol = 0.9, margins = c(8, 8),
          trace = "none", density.info = "none",
          main = "Single-term presidents")

par(mar=c(1,1,1,1))
topic.summary=tbl_df(corpus.list.df)%>%
  filter(Type == "Multi-Term")%>%
  select(File, 23:37)%>%
  group_by(File)%>%
  summarise_all(funs(mean))
topic.summary=as.data.frame(topic.summary)
rownames(topic.summary)=topic.summary[,1]

heatmap.2(as.matrix(topic.summary[,topic.plot+1]), 
          scale = "column", 
          col = rich.colors(100),
          cexRow = 0.9, cexCol = 0.9, margins = c(8, 8),
          trace = "none", density.info = "none",
          main = "Multi-term presidents")
```

Some presidents really focused on a topic, while others spread across multiple topics.
Recent multi-term presidents talked about freedom, hope, new Americans (immigration?), and world peace.

Recent single-term presidents, including Donald Trump, also talked about similar topics, but they also talked about "will-one-now" which contains words like believ, make, race, good, and give.

Maybe there are more differecences between parties?
###

```{r}
par(mfrow=c(2, 2))

speech.df=tbl_df(corpus.list.df)%>%
  filter(Type == "Single-Term")%>%
  select(Party, 23:37)%>%group_by(Party)%>%
  summarise_all(funs(mean))
speech.df=as.matrix(speech.df)
speech.df[,-1]=replace(speech.df[,-1], speech.df[,-1]<1/15, 0.001)
party = c("Democratic","Republican")
speech.df = data.frame(party,speech.df[,-1],row.names = NULL)
speech.df.m = melt(speech.df,id.vars='party')
speech.df.m.ordered <- speech.df.m[order(speech.df.m$value, decreasing = TRUE),]
speech.df.m.ordered$party <- factor(speech.df.m.ordered$party, 
                                    levels = c("Republican","Democratic"))

ggplot(data = subset(speech.df.m.ordered)[speech.df.m.ordered$party=="Republican",],
          aes(x=factor(1),y=value,fill = variable)) +
  geom_bar(width = 1, stat="identity") + coord_polar(theta="y") +
  xlab("Republican") + ylab("Single Term") +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid  = element_blank()
  )

ggplot(data = subset(speech.df.m.ordered)[speech.df.m.ordered$party=="Democratic",],
          aes(x=factor(1),y=value,fill = variable)) +
  geom_bar(width = 1, stat="identity") + coord_polar(theta="y") +
  xlab("Democratic") + ylab("Single Term") + 
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid  = element_blank())


speech.df=tbl_df(corpus.list.df)%>%
  filter(Type == "Multi-Term")%>%
  select(Party, 23:37)%>%group_by(Party)%>%
  summarise_all(funs(mean))
speech.df=as.matrix(speech.df)
speech.df[,-1]=replace(speech.df[,-1], speech.df[,-1]<1/15, 0.001)
party = c("Democratic","Republican")
speech.df = data.frame(party,speech.df[,-1],row.names = NULL)
speech.df.m = melt(speech.df,id.vars='party')
speech.df.m.ordered <- speech.df.m[order(speech.df.m$value, decreasing = TRUE),]
speech.df.m.ordered$party <- factor(speech.df.m.ordered$party, 
                                    levels = c("Republican","Democratic"))


ggplot(data = subset(speech.df.m.ordered)[speech.df.m.ordered$party=="Republican",],
          aes(x=factor(1),y=value,fill = variable)) +
  geom_bar(width = 1, stat="identity") + coord_polar(theta="y") +
  xlab("Republican") + ylab("Multi Term") +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid  = element_blank()
  )

ggplot(data = subset(speech.df.m.ordered)[speech.df.m.ordered$party=="Democratic",],
          aes(x=factor(1),y=value,fill = variable)) +
  geom_bar(width = 1, stat="identity") + coord_polar(theta="y") +
  xlab("Democratic") + ylab("Multi Term") + 
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid  = element_blank())
  
```

Republicans seem to focus on "will-one-now", but there are some differences between single- vs multi-term presidents among them.

Democrats seem to focus on "peopl-without-civil" and "great-never-countri" if they are single term presidents, but multi-term presidents focus on "can-must-peopl" and "peopl-without-civil".

As the next step of this analysis, it'd be interesting to compare Donald Trump's speech to those by multi-term presidents and single-term presindents, and see if we can predict he will be elected for the second term.
